---
title: "Data Processing | 2014, Yellow Cab Trip Record Data"
author: "Group No. 2"
date: "4/22/2019"
output: html_document
---

****** USE A GIT IGNORE FILE BEFORE COMMITTING TO GIT******
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Set the working directory ------
setwd(Users/jntay/Box Sync/Data Science Project - Trip Record Data)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(glimpse)
require(readr)
require(lubridate)
require(recipes)
```


## Reading in data and processing ----
Reading in and processing Trip Record Data from New York City's Taxi and Limousine Commission. Data is saved by year, month, and vehicle type. We will download data from 2014-18 for each vehicle type and then transform/wrangle the data for analysis.

read in data,
mutate to reduce the size
delete the original file
any other calcs
save new file

Yellow cab data for 2014 had different variable names which required slight changes to the code.


Reading in one file at a time
```{r}
yellow_data <- read_csv("yellow_tripdata_2016-07.csv")

```

```{r}
# Clean data

yellow <- yellow_data %>%  
  select(pickup_date = pickup_datetime,  # selecting variables we are interested in and renaming
         dropoff_date = dropoff_datetime,
         passenger_count,
         trip_distance,
         payment_type,
         total_fare = total_amount
    ) %>%
  
  add_column(vehicle_type= "yellow") # add vehicle type variable


# mutate variables

yellow_mutate <- yellow %>%
  mutate(pickup_date = round_date(pickup_date, unit = "day"),  #round date-times to day only, drop time
         dropoff_date = round_date(dropoff_date, unit = "day"),
         month = month(pickup_date), # create month variable
         year = year(pickup_date), # create year variable
         payment_type = as.factor(payment_type)) # coerce to factor variable
str(yellow_mutate)
```

We notice that `payment_type` is coded differently for 2014, which will require a slight change to the code
```{r}
summary(yellow_mutate$payment_type)
```


```{r}
# Dummy out the payment type categories using the recipes package
dummy_payment <- recipe(pickup_date~.,yellow_mutate) %>%
  step_dummy(payment_type, one_hot = TRUE) %>%  
prep()

yellow_clean <- bake(dummy_payment, yellow_mutate) %>%
  select(pickup_date,
         dropoff_date,
         month,
         year,
         passenger_count,
         trip_distance,
         total_fare,
         vehicle_type,
         payment_credit = payment_type_CRD,
         payment_cash = payment_type_CSH,
         payment_no_charge = payment_type_NOC,
         payment_dispute = payment_type_DIS,
         payment_unknown = payment_type_UNK
         )
str(yellow_clean)
```



**Code for dates not already in date-time format**
yellow <- yellow %>%
  mutate(pickup_date = mdy_hm(`pickup_date`),    # convert strings to date-times
         pickup_date = as_date(pickup_date),   # mutate date-times to date only
         dropoff_date = mdy_hm(`dropoff_date`),
         dropoff_date = as_date(dropoff_date))

mutate(pickup_date = as.Date(pickup_date, format ="%Y-%m-%d"),
         dropoff_date = as.Date(pickup_date, format ="%Y-%m-%d"))
         
**Code for `pickup_date` and `dropoff_date`already in date-time format**
(pickup_date = round_date(pickup_date, unit = "day"),  #round date-times to day only, drop time
         dropoff_date = round_date(dropoff_date, unit = "day"),


```{r}
# create function to calculate mode for payment type factor variables
mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
```


```{r}
#aggregate data up to a month and summarise
 yellow_monthly <- yellow_clean %>%
 mutate(pickup_date = round_date(pickup_date, unit = "month"),
       dropoff_date = round_date(dropoff_date, unit = "month")) %>%
  group_by(month, year, vehicle_type) %>%
  summarise(mean_passenger_count= mean(passenger_count), 
            payment_credit = sum(payment_credit),
            payment_cash = sum(payment_cash),
            payment_no_charge = sum(payment_no_charge),
            payment_dispute = sum(payment_dispute),
            mean_trip_distance = mean(trip_distance),
            mean_total_fare = mean(total_fare),
            trip_n = n()) %>% 
  na.omit() %>%
  filter(year==2014 & month==01) # filtering out any erroneous dates


```

Save aggregated data to csv file
```{r}
write_csv(yellow_monthly, "~/Box Sync/Data Science Project - Trip Record Data/Yellow Clean/clean-yellow-2014-01.csv")

```

Repeat for each file/month