---
title: "Data Science Project: New York City Taxi Market"
author: "Group No. 2: McCourt GitHubers "
        "Yihan Wang", "Natalia Gonzalez-Alarcon", "Jamie Taylor"
date: "March 30, 2019"
output: html_document
---

#Set working directory
setwd(Users/jntay/Box Sync/Data Science Project - Trip Record Data)

USE A GIT IGNORE FILE BEFORE COMMITTING TO GIT

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(glimpse)
require(readr)
require(lubridate)
require(recipes)
```

## Part 1 -------------------------------------------------------

## Reading in data and processing --------------------------------
Reading in and processing Trip Record Data from New York City's Taxi and Limousine Commission. Data is saved by year, month, and vehicle type. We will download data from 2014-18 for each vehicle type and then transform/wrangle the data for analysis.

read in data,
mutate to reduce the size
delete the original file
any other calcs
save new file


Reading in one file at a time to mutate
```{r}
# Read in one file ----------------

yellow_data <- read_csv("yellow/yellow_tripdata_2018-02.csv")

```

```{r}
# Clean data

yellow <- yellow_data %>%  
  select(pickup_date = tpep_pickup_datetime,  # selecting variables we are interested in and renaming
         dropoff_date = tpep_dropoff_datetime,
         passenger_count,
         trip_distance,
         payment_type,
         total_fare = total_amount,
         pickup_location = PULocationID,
         dropoff_location = DOLocationID
    ) %>%
  
  add_column(vehicle_type= "yellow") # add vehicle type variable
```

```{r}
# mutate variables

yellow_mutate <- yellow %>%
  mutate(pickup_date = round_date(pickup_date, unit = "day"),  #round date-times to day only, drop time
         dropoff_date = round_date(dropoff_date, unit = "day"),
         month = month(pickup_date), # create month variable
         year = year(pickup_date), # create year variable
         payment_type = as.factor(payment_type), # coerce to factor variable
         trip = paste(pickup_location, dropoff_location, sep = "-", collapse = NULL)) %>% # convert PU and DO location to strings and concatenate
  select(-pickup_location, -dropoff_location) # drop pickup and dropoff location variables
str(yellow_mutate)

```

```{r}
# Dummy out the payment type categories using the recipes package
dummy_payment <- recipe(pickup_date~.,yellow_mutate) %>%
  step_dummy(payment_type, one_hot = TRUE) %>%  
prep()

yellow_clean <- bake(dummy_payment, yellow_mutate) %>%
  select(pickup_date,
         dropoff_date,
         month,
         year,
         passenger_count,
         trip_distance,
         total_fare,
         vehicle_type,
         trip,
         payment_credit = payment_type_X1,
         payment_cash = payment_type_X2,
         payment_no_charge = payment_type_X3,
         payment_dispute = payment_type_X4
         )
str(yellow_clean)
```

Some variables did not read in as date-time variables
**Code for dates not already in date-time format**
```{r}
# yellow <- yellow %>%
#  mutate(pickup_date = mdy_hm(`pickup_date`),    # convert strings to date-times
#         pickup_date = as_date(pickup_date),   # mutate date-times to date only
#         dropoff_date = mdy_hm(`dropoff_date`),
#         dropoff_date = as_date(dropoff_date))
```

Create function to calculate mode for payment type factor variables 
```{r}
mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
```

Aggregate data up to the month level and summarise
```{r}
 yellow_monthly <- yellow_clean %>%
 mutate(pickup_date = round_date(pickup_date, unit = "month"),
       dropoff_date = round_date(dropoff_date, unit = "month")) %>%
  group_by(month, year, vehicle_type) %>%
  summarise(mean_passenger_count= mean(passenger_count), 
            payment_credit = sum(payment_credit),
            payment_cash = sum(payment_cash),
            payment_no_charge = sum(payment_no_charge),
            payment_dispute = sum(payment_dispute),
            mean_trip_distance = mean(trip_distance),
            mean_total_fare = mean(total_fare),
            trip = mode(trip),
            trip_n = n()) %>% 
  na.omit() %>%
  filter(year==2018 & month==1)

```

Save new monthly data to csv file
```{r}
write_csv(yellow_monthly, "clean-yellow-jan-2018.csv")
```

Repeat steps above for each file